Resonant Scattering in a Uniform Sphere with Large Optical Depth

### 2021-07-16 ms.tex
Adding gamma_nm plot and beginnings of explanation

The solution of the radiative transfer equation for resonant scattering of Lyman $\alpha$ photons for a uniform sphere of constant gas density is found in the limit of large line center optical depth. A monochromatic source of photons is assumed at the center of the sphere, and the intensity within the sphere and at the surface is found with the Eddington approximation for the angular dependence. The solution can be represented a sum of two terms: a semi-analytic solution of the homogeneous equation which is required for the mean intensity $J=0$ at the surface, and a solution of the homogeneous equation which enforces the correct, zero-ingoing intensity boundary conditions at the surface. It is the compact analytic form as well as the piece accounting for the correct, frequency-dependent boundary condition that is novel in this study. The analytic solution is compared to numerical solutions with the Monte-Carlo method, which are valid at arbitrary optical depth. Discrepancies in the spectra of previous work are quantified, and it is shown that the surface boundary condition derived here corrects these discrepancies. Separately, the time-dependence of the homogeneous equation is calculated using an eigenfunction expansion and solved numerically to characterize the time-to-escape of photons initialized at the center of the sphere, including the effects of frequency redistribution. This distribution can be sampled to accelerate Monte Carlo radiation transfer calculations at very large optical depths, obtaining intensities and radiation forces on cell faces at a fraction of the computational cost.

@@

@@

The eigenfunction expansion method is an effective characterization of the time-to-escape, yet not a fast enough computational method to use as a distribution for sampling. However, the behavior of the solution is well-understood since the late-time distribution is simply an exponential falloff in photon escape probability. The rate constant of the function is the lowest-order eigenfrequency, $\gamma_{00}$, and its scale is determined by the coefficient $P_{00}$ as in Eq. \ref{eq:pnmsoln}. Thus, an approximate ``fitting function'' that captures both the peak of the escape time distribution and the exponential falloff is
\be \label{eq:fitting_function}
P(t) = \exp{\left[-\left(\frac{t_{\rm diff}}{t}\right)^2\right]} \times \gamma_{00} P_{00} e^{-\gamma_{00}t}.
\ee
The first term represents the early-time distribution, which then transitions to an exponential falloff past a point $ct/R = ct_{\rm diff}/R = (a\tau_0)^{1/3}$, where $t_{\rm diff}$ is a characteristic diffusion timescale. This is shown in Fig. \ref{fig:escape_time} for two different line center optical depths. It is seen that the fitting function matches well with the late-time behavior of the eigenfunction solution, indicating that the tail of the escape time distribution can be modeled effectively by simple exponential decay. The only remaining discrepancy between the tail of the distribution and the Monte Carlo data is due to core scattering, which is not accurately modeled by the eigenfunction solution as previously discussed. A cumulative distribution function can be created from Eq. \ref{eq:fitting_function} using a simple rejection method, which allows the function to be sampled rapidly.

@@

The time-dependence of the problem is solved under a $J=0$ boundary condition to characterize the distribution of photon escape times. The differential equation cannot be solved except with a semi-analytic approach, utilizing an expansion in both space and photon frequency. By matching boundary conditions on either side of the photon source frequency, the flux at the surface of the sphere is found as a function of time and photon frequency, expressed as a sum over many eigenmodes $n$ and $m$---spatial and frequency modes, respectively. Calculating additional spatial eigenmodes provide a more accurate model of the physics nearer to line center, but convergence is slow due to each eigenmode's weak dependence on $n$. Additional frequency eigenmodes introduce fast-oscillating terms that improve the accuracy of the sum, as their contributions cancel with components of lower-order terms to better represent the true solution. Integrating the solution over time produces a spectrum that is shown to agree well with the steady-state calculations in Section \ref{sec:steadystate}, provided a sufficient number of terms in the sum. Integrating the solution over frequency leads to a distribution of photon escape times, which can be compared directly with Monte Carlo simulations. The sum over eigenmodes produces an escape-time distribution that broadly captures the behavior shown by Monte Carlo data---a rise at early times, transitioning to exponential decay in the tail of the distribution. It is expected that the accuracy of the rate constant for the tail of the distribution is limited by the number of spatial terms included in the sum over eigenmodes, as the Doppler core can trap photons at low mean free paths until they diffuse outward in frequency, biasing the distribution toward later times. This physics is not modeled by our solution for two reasons: 1) our calculations ignored the Gaussian component of the Voigt line profile, leaving the Lorentzian piece which is accurate only in the line wing, and 2) knowing the core is not modeled accurately, we do not include a large enough number of spatial eigenmodes in the sum to resolve it. However, a fitting function dependent on parameters $a$ and $\tau_0$ is found that adequately represents the escape time distribution within these constraints. When calculated as a cumulative distribution function using a rejection method, the fit can be used to sample photon escape times directly rather than computing their travel times by the Monte Carlo method. As will be shown in future work, these calculations can be implemented as an acceleration scheme for Monte Carlo resonance-line radiation transfer at optical depths that would nominally be prohibitive in computational cost.

>>>>

The solution of the radiative transfer equation for resonant scattering of Lyman $\alpha$ photons for a uniform sphere of constant gas density is found in the limit of large line center optical depth. A monochromatic source of photons is assumed at the center of the sphere, and the intensity within the sphere and at the surface is found with the Eddington approximation for the angular dependence. The solution can be represented a sum of two terms: a semi-analytic solution of the homogeneous equation which is required for the mean intensity $J=0$ at the surface, and a solution of the homogeneous equation which enforces the correct, zero-ingoing intensity boundary conditions at the surface. It is the compact analytic form as well as the piece accounting for the correct, frequency-dependent boundary condition that is novel in this study. The analytic solution is compared to numerical solutions with the Monte-Carlo method, which are valid at arbitrary optical depth. It is shown that the scale of corrective term depends on the line center optical depth, accounting for discrepancies in the spectra of previous studies that used a $J=0$ boundary condition. Separately, the time-dependence of the diffusion equation is expressed using an eigenfunction expansion and solved numerically to characterize the time-to-escape of photons initialized at the center of the sphere. It is shown that the semi-analytic calculation agrees well with the Monte Carlo escape time distribution, converging for large line center optical depths around $\tau_0=10^7$. A fitting function for this distribution is found which can be sampled to accelerate Monte Carlo radiation transfer calculations at very large optical depths.

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{gamma_nm.pdf}
    \caption{The scaling of numerically-obtained resonant frequencies as compared with the analytic expression in Eq. \ref{eq:gamma_nm}.}
    \label{fig:gamma_nm}
\end{figure}

TODO: The values of the $\gamma_{nm}$ resonant frequencies are calculable analytically. Their scale depends on $m$, $n$, and the factor $(a\tau_0)^{1/3}$ according to 
\be \label{eq:gamma_nm}
2**(-1/3) * np.pi**(13/6)*n**(4/3)*(m-7/8)**(2/3)*fc.clight/p.radius/(p.a * p.tau0)**(1/3.) 
\ee
The size of additional eigenfrequencies scales as a powerlaw in $m$, yet as is seen in Fig. \ref{fig:gamma_nm} the scaling is weak, requiring up to 500 $m$ to reduce the scale of $\gamma_{nm}^{\ \ -1}$ by two orders of magnitude.

@@

The eigenfunction expansion method is an effective characterization of the time-to-escape, yet not a fast enough computational method to use as a distribution for sampling. However, the behavior of the solution is well-understood since the late-time distribution is simply an exponential falloff in photon escape probability. The rate constant of the function is the lowest-order eigenfrequency, $\gamma_{00}$, and its scale is determined by the coefficient $P_{00}$ as in Eq. \ref{eq:pnmsoln}. Thus, an approximate ``fitting function'' that captures both the peak of the escape time distribution and the exponential falloff is
\be \label{eq:fitting_function}
P(t) = \exp{\left[-\left(\frac{t_{\rm diff}}{t}\right)^2\right]} \times \gamma_{00} P_{00} e^{-\gamma_{00}t}.
\ee
The first term represents the early-time distribution, which then transitions to an exponential falloff past a point $ct/R = ct_{\rm diff}/R = (a\tau_0)^{1/3}$, where $t_{\rm diff}$ is a characteristic diffusion timescale. This is shown in Fig. \ref{fig:escape_time} for two different line center optical depths. It is seen that the fitting function matches well with the late-time behavior of the eigenfunction solution, indicating that the tail of the escape time distribution can be modeled effectively by simple exponential decay. The remaining discrepancy between the tail of the distribution and the Monte Carlo data is due to core scattering, which is not accurately modeled by the eigenfunction solution. A cumulative distribution function can be created from Eq. \ref{eq:fitting_function} using a simple rejection method, which allows the function to be sampled rapidly.

@@

The time-dependence of the problem is solved under a $J=0$ boundary condition to characterize the distribution of photon escape times. The differential equation cannot be solved except with a semi-analytic approach, utilizing an expansion in both space and photon frequency. By matching boundary conditions on either side of the photon source frequency, the flux at the surface of the sphere is found as a function of time and photon frequency, expressed as a sum over many eigenmodes $n$ and $m$---spatial and frequency modes, respectively. Calculating additional spatial eigenmodes provides a more accurate model of the physics nearer to line center, but convergence is slow due to each eigenmode's weak dependence on $n$. Additional frequency eigenmodes introduce fast-oscillating terms that improve the accuracy of the sum, as their contributions cancel with components of lower-order terms to better represent the true solution. Integrating the solution over time produces a spectrum that is shown to agree well with the steady-state calculations in Section \ref{sec:steadystate}, provided a sufficient number of terms in the sum. Integrating the solution over frequency leads to a distribution of photon escape times, which can be compared directly with Monte Carlo simulations. The sum over eigenmodes produces an escape-time distribution that broadly captures the behavior shown by Monte Carlo data---a rise at early times, transitioning to exponential decay in the tail of the distribution. It is expected that the accuracy of the rate constant for the tail of the distribution is limited by the number of spatial terms included in the sum over eigenmodes, as the Doppler core can trap photons at high optical depths until they diffuse outward in frequency, biasing the distribution toward later times. This physics is not modeled by our solution for two reasons: 1) our calculations ignored the Gaussian component of the Voigt line profile, leaving the Lorentzian piece which is accurate only in the line wing, and 2) knowing the core is not modeled accurately, we do not include a large enough number of spatial eigenmodes in the sum to resolve it. However, a fitting function dependent on parameters $a$ and $\tau_0$ is found that adequately represents the escape time distribution within these constraints. When calculated as a cumulative distribution function using a rejection method, the fit can be used to sample photon escape times directly rather than computing their travel times by the Monte Carlo method. As will be shown in future work, these calculations can be implemented as an acceleration scheme for Monte Carlo resonance-line radiation transfer at optical depths that would nominally be computationally prohibitive.
